---
title: "Introduction to SPDtrace: Efficient Learning of Differential Networks"
author: "Mojtaba Nikahd, Seyed Abolfazl Motahari"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to SPDtrace}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Introduction

The **SPDtrace** package implements the SPD-trace method for efficient learning of differential networks in multi-source non-paranormal graphical models. This package provides tools for differential network inference, simulation studies, and real-world applications in genomics and other high-dimensional data analysis.

## Installation

```{r install, eval = FALSE}
# Install from CRAN (when available)
install.packages("SPDtrace")

# Or install from GitHub
# devtools::install_github("yourusername/SPDtrace")
```

## Quick Start

### Basic Usage

The main function `SPDtrace()` provides a simple interface for differential network inference:

```{r quick_start, eval = FALSE}
library(SPDtrace)

# Generate example data
set.seed(123)
n <- 100
p <- 20

# Create two covariance matrices
Sigma_A <- diag(p) + 0.3 * (matrix(runif(p^2), p, p) > 0.8)
Sigma_A <- (Sigma_A + t(Sigma_A)) / 2
diag(Sigma_A) <- 1

Sigma_B <- Sigma_A + 0.2 * (matrix(runif(p^2), p, p) > 0.9)
Sigma_B <- (Sigma_B + t(Sigma_B)) / 2
diag(Sigma_B) <- 1

# Run SPD-trace
result <- SPDtrace(CovA = Sigma_A, CovB = Sigma_B, sparsityLevel = 50)

# View results
print(result)
summary(result)
```

### Understanding the Output

The `SPDtrace()` function returns an object of class `SPDtrace_result` containing:

- `solution_path`: List of solution path objects with knots and active sets
- `differential_network`: Adjacency matrix of the inferred differential network
- `lambda_sequence`: Sequence of regularization parameters
- `method`: Method used for inference
- `call`: Function call

## Core Functions

### 1. Main Inference Function

```{r main_function, eval = FALSE}
# Basic usage
result <- SPDtrace(CovA, CovB, sparsityLevel = 30)

# With different methods
result_spd <- SPDtrace(CovA, CovB, sparsityLevel = 30, method = "SPDtrace")
result_cross <- SPDtrace(CovA, CovB, sparsityLevel = 30, method = "CrossFDTL")
result_dtrace <- SPDtrace(CovA, CovB, sparsityLevel = 30, method = "DTrace")
```

### 2. Utility Functions

The package provides several utility functions for data preprocessing and analysis:

```{r utilities, eval = FALSE}
# Convert Kendall's tau to Pearson correlation
tau_matrix <- matrix(c(1, 0.5, 0.3, 0.5, 1, 0.4, 0.3, 0.4, 1), 3, 3)
pearson_matrix <- kendall_tau_matrix_to_pearson_correlation_matrix(tau_matrix)

# Ensure positive semi-definiteness
A_psd <- positive_semi_definite_maker(A)

# Estimate correlation from multiple datasets
cor_matrix <- weighted_rank_based_pearson_correlation_estimator(datasets)
```

### 3. Simulation Functions

Run simulation studies to evaluate method performance:

```{r simulation, eval = FALSE}
# First scenario: Performance comparison
results_scenario1 <- run_first_scenario(
  n_samples = 100, 
  n_variables = 20, 
  n_replicates = 10
)

# Second scenario: Robustness testing
results_scenario2 <- run_second_scenario(
  n_samples = 100, 
  n_variables = 20, 
  n_replicates = 10
)
```

## Real-World Application: Ovarian Cancer

The package includes real data from The Cancer Genome Atlas (TCGA) for ovarian cancer analysis:

```{r real_data, eval = FALSE}
# Load ovarian cancer data
data("ov_gene_exp")
data("ov_pi3k_akt_mtor")
data("ov_wdtrace_mat")
data("platinum_resistant_genes")

# Analyze differential networks between platinum-sensitive and resistant tumors
# (This would be implemented in a separate analysis script)
```

## Advanced Features

### Model Selection

The package supports various model selection strategies:

```{r model_selection, eval = FALSE}
# Use instability evaluation for model selection
instability_results <- instability_evaluator_of_solution_paths(
  solution_paths, 
  number_of_nodes
)

# Fisher's test for biological validation
enrichment_result <- fisher_test(
  all_genes, 
  hub_genes, 
  functional_genes
)
```

### Custom Methods

You can implement custom differential network inference methods:

```{r custom_methods, eval = FALSE}
# The package is designed to be extensible
# You can add new methods by implementing the required interface
```

## Performance Considerations

### Memory Usage

- For large networks (p > 1000), consider using sparse matrix representations
- The C++ implementation provides efficient memory management
- Use appropriate sparsity levels to control computational complexity

### Parallelization

- Simulation studies can be parallelized using the `parallel` package
- Consider using `future` package for advanced parallel processing

## Troubleshooting

### Common Issues

1. **C++ compilation errors**: Ensure Rtools (Windows) or Xcode (macOS) is properly installed
2. **Memory issues**: Increase R memory limit or use garbage collection
3. **Package dependencies**: Install all required packages from CRAN

### Getting Help

- Check the package documentation: `?SPDtrace`
- View function examples: `example(SPDtrace)`
- Report issues on the GitHub repository

## References

Nikahd, M., & Motahari, S. A. (2025). Efficient Learning of Differential Networks in Multi-Source Non-Paranormal Graphical Models. arXiv preprint arXiv:2410.02496.

## Session Info

```{r session_info, echo = FALSE}
sessionInfo()
```
